---
title: "paremeters_tuning"
author: "Grigorii Nos"
date: "2024-04-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dev='CairoPNG', fig.width=10, fig.height=6, dpi=300)
```

## Parameters tuning for pMHC-TCR assigments

In this vignette we will explore how different parameters affect TCR-pMHC assigment
and how to tune them. 

```{r load_packages, echo=FALSE, message=FALSE, warning=FALSE}
library(Seurat)
library(pmhcDenoiseR)
library(uniformly)
library(tidyverse)
library(ggplot2)
library(ComplexHeatmap)
```

Let's load downsampled 10x dataset, that includes 51273 from 1000 clones. 1000 clones
were downsampled from a bigger dataset to save us some computational time. 

```{r}
data('10x_dataset')

tenx.data@assays$pMHC@counts[,1:5] %>% head()
```

let's asses the noise level for each pMHC and determine from what Entropy value
we want to skip pMHC features from TCR-pMHC assigment, if the clone is small in size

```{r}
hm_noise <- pmhc_heatmap(object = tenx.data, slot='counts', clones = downsampled_clones, clones_order = NULL,
                         highlight_pmhc.tcr=F, hm_breaks = c(0, 5, 10), split_cols = F,
                         show_row_names = T, clean_mat = F, max_cols = 10000)

left_ann_entropy <- rowAnnotation(
  entropy = anno_barplot(tenx.data@misc$noise_score$entropy,
                         width = unit(1, "cm"), 
                         gp = gpar(fill = "blue"),
                         axis_param = list(direction = "reverse")))

left_ann_entropy + hm_noise

```
The appropriate threshold seems to be around 0.7

This data is already pre-scaled (without downsampling using 208589 gems) 
and pre-smoothed. Here we will explore delta, major parameter for pmhc_assign.

Let see what happens with delta=1

```{r}
tenx.data <- assign_pmhc(object = tenx.data, cl_size_thresh = 3, 
                         entropy_thresh = 0.7, 
                         assign_small_clones = T,
                         delta_threshold = 1)

pmhc_heatmap(object = tenx.data, stop_large_highlighting = F,
             slot='counts', clones = downsampled_clones, clones_order = NULL,
             highlight_pmhc.tcr=T, hm_breaks = c(-5, 0, 5), split_cols = F,
             show_row_names = T, clean_mat = F, max_cols = 10000)
```




