---
title: "paremeters_tuning"
author: "Grigorii Nos"
date: "2024-04-08"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(dev='CairoPNG', fig.width=10, fig.height=6, dpi=300)
```

## R Markdown

In this vignette you can see how parameters of pmhcDenoiseR can be optimised for
your specific dataset, and how to handle more complex scenarios than in viral_screen
example. 

```{r load_packages, echo=FALSE, message=FALSE, warning=FALSE}
library(Seurat)
library(pmhcDenoiseR)
library(uniformly)
library(tidyverse)
library(ggplot2)
library(ComplexHeatmap)
```

### load the data

Here is a 10x dataset with 49 different pMHC multimers from downsamples 1000 clones,
among these 1000 clones are the ones with different specificities and clone sizes.

```{r}
data('10x_dataset')

tenx.data@assays$pMHC@counts[,1:5] %>% head()
```


This dataset is already pre-scaled (without downsampling, on a larger set of cells (208589))
and pre smoothed. Here we want to explore a major parameter that changes the pmhc assigment,
delta value. The default is 1, let's start with that.


```{r}
object <- assign_pmhc(object, cl_size_thresh = 3, entropy_thresh = 0.7, delta_threshold = 0.8)

pmhc_heatmap(object = obj, stop_large_highlighting = F,
             slot='counts', clones = downsampled_clones, clones_order = NULL,
             highlight_pmhc.tcr=T, hm_breaks = c(0, 5, 10), split_cols = F,
             show_row_names = T, clean_mat = F, max_cols = 16000)
```





