---
title: "ITRAP2 vignette"
author: "Grigorii Nos"
date: "2025-07-15"
output:
  rmarkdown::github_document:
    toc: true
    toc_depth: 2
    html_preview: false
    md_extensions: +tex_math_dollars+raw_tex
  rmarkdown::html_vignette: default   # optional: for R CMD build/check
vignette: >
  %\VignetteIndexEntry{ITRAP2 vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, 
                      warning = FALSE, fig.align = "center", 
                      dev='CairoPNG', fig.width=10, fig.height=6, dpi=300)
```

## R Markdown

ITRAP2 is a data denoising workflow for single-cell pMHC datasets. The workflow also assigns the 
pMHC specificities to individual clonotypes. ITRAP2 works with coupled single-cell pMHC counts with TCR clonotypes.

### load libraries
```{r load_packages, message=FALSE, warning=FALSE}
library(Seurat)
library(ITRAP2)
library(uniformly)
library(tidyverse)
library(ggplot2)
library(ComplexHeatmap)
```

### load the data

we generated a dataset with PBMC T cells from healthy donors, that were screen for the recognition 100 pMHC panel of common chronic viral epitopes, the data is stored as a Seurat object, as the whole package is build around Seurat class

```{r}
data('viral_screen')

viral_screen@assays$pMHC@counts[,1:5] %>% head()
```

the gene expression part of this object is deleted here to save the space, in case you want to explore the data with gene expression, you are welcome to download it from ArrayExpress repository. https://www.ebi.ac.uk/biostudies/arrayexpress/studies/E-MTAB-13758?key=686eea8a-f64c-4c66-a859-9ca937cfa767

you can see that features aren't directly names as pMHC, sinse in some experients the same barcode can by applied to different pMHC in different donors, so we store this data.frame for pMHC annotation

```{r pmhc_annotation}

viral_screen@misc$pmhc %>% head()       
```

it will help to assign specificities in the @meta.data later on.

We also expect you to store your clonotype information for each gem in @meta.data, we use `clone_id` column name, sinse we used immcantation pipeline https://immcantation.readthedocs.io/en/stable/ to process TCR sequencing data.

```{r}
viral_screen@meta.data$clone_id %>% head(10)
```

# visualise raw data

Lets first look at the pMHC distribution of the raw pMHC count within big clviral_screen, we expect a consistent signal for a specific pMHC within some clviral_screen, before labeling the clone with the specificity, but we also expect a substantial amount of noise.

```{r}
# we subset gems from 1 donor for simplicity, bc=buffy coat, BC390 is a donor ID
obj_sub <- subset(viral_screen, BC == 'BC390') 

# we take obly 5+ gems clviral_screen  
clones <- obj_sub$clone_id[obj_sub$clone_size > 5] %>% table() %>% sort(decreasing = T) %>% names()

# and we plot the heatmap, sorted by clonotypes
pmhc_heatmap(object = obj_sub, slot = 'counts', hm_breaks = c(0, 5, 10), 
             rowm.fonts = 5, split_cols = T,
             clones=clones, show_row_names = T, clean_mat = F)
```
we can see that some clone (grouped columns) have noisy, but visible consistent signals some pMHCs, we can further visualise a single clone in more details

```{r }
pmhc_dist_per_clone(object = viral_screen, clone = '3487_2473', aggr_threshold = 10, slot = 'counts')
```
Look at this clone and pMHC counts distributions, on the left we can see that two B0801 pMHC have the highest number of gems, that aren't zero. We can as well calculate a metric called clonal pmhc concordance, introduced in Polvsen et al 2023, elife, that calculates a proportion of gems in the clone, whose highest value of pMHC count is in this pMHC. So we can reasonably assume that this clone is B0801 specific, but data isnot clean enough to have any automate way of assigning it. 

Next we want to estimate the level of noisyness of each pMHC, we can do that by calculating an entropy ( https://en.wikipedia.org/wiki/Entropy_(information_theory) ) for each pMHC, the pMHCs with highest entropy are expected to be the noisiest

```{r echo=TRUE, message=FALSE, warning=FALSE}
viral_screen <- score_pmhc_noise(viral_screen, how = 'per_clone', downsample_rate = .1, verbose=F)

hm_noise <- pmhc_heatmap(object = obj_sub, slot = 'counts', hm_breaks = c(0, 5, 10), 
                         rowm.fonts = 5, split_cols = F,
                         clones=clones, show_row_names = T, clean_mat = F)

left_ann_entropy <- rowAnnotation(
  entropy = anno_barplot(viral_screen@misc$noise_score$entropy,
                         width = unit(1, "cm"), 
                         gp = gpar(fill = "blue"),
                         axis_param = list(direction = "reverse")))

left_ann_entropy + hm_noise
```

You can see that the pMHCs with the highest entropy level procude the most noise, we will use that metric to calculate confidence scores for each pMHC-TCR pairs, where higher entropy lowers our confidence. 

## scaling

So in next steps we can start to normalize and denoise the data, we apply specific way of mean centering and sd scaling, where means and sds are calculated, not taking outlines (|z|>3) into account. this approach is used, instead of typically used median centering and mad scaling, since in sparse single-cell datasets, medians and mads are often zeros.

We prefer to perform scaling on the whole dataset for better mean and sd estimation.
```{r message=FALSE, warning=FALSE}
viral_screen <- ScaleDataNoOutliers(viral_screen, verbose = F) # records scaled matrix into @assays$pMHC@data

obj_sub <- subset(viral_screen, BC == 'BC390') 

pmhc_heatmap(object = obj_sub, slot = 'data', hm_breaks = c(-3, 0, 10),
             split_cols = T, rowm.fonts = 5, clones = clones, 
             show_row_names = T, clean_mat = F)
```

Explore a single clone within scaled UMI counts space

```{r}
pmhc_dist_per_clone(object = obj_sub, clone = '3487_2473', aggr_threshold = 10, slot = 'data')
```

We can clearly see more clear signal, that we can further improve in the next step, which is smoothing each pMHC within each expanded clone, using local regression (loess). 

## smoothing

```{r message=FALSE, warning=FALSE}
obj_sub <- smooth_pmhc(object = obj_sub, assay = 'pMHC', 
                       cap_upper_quantiles = T, # clip huge outliers, turning them into the replacement quantile, default=T
                       replacement_q = .85, # which quantile to use for the clipping
                       replace_ones = T  # replace 1 UMI count with zero pre-smoothing
                       )
```

and next we can start pairing TCRs with pMHCs, not for a small clone <3 gems we will not take into account pMHC with entropy higher than 0.7. This function will also calculate confidence score for each pMHC pair. Green frames highlight the assigned specificities

```{r message=FALSE, warning=FALSE}
obj_sub <- assign_pmhc(object = obj_sub, assay = 'pMHC', slot = 'scale.data', 
                       assignment = 'rosner', # use extreme alternatively
                       rosner_alpha=.001, verbose = F, # from which alpha to assign the pMHC 
                       rosner_pval_dist = 'normal', 
                       remove_for_params=2, params='remove_topx', pseudobulk_fun=median,
                       adjust_permutation = F, adjust_test_within_clone = T, 
                       padj_method = 'BH', assign_small_clones = T)

```

We recommend adjusting this parameters for assign_pmhc, checking how your assignment look on the pMHC heatmap, in some datasets certain parameters fail, which you can only see from visualizing your assignments. For more cross-reactivity heavy datasets, try using extreme, instead of rosner assignments. If you feel like a lot of potential specificities are missed, try to change alpha in assign_pmhc. 

```{r}
pmhc_heatmap(object = obj_sub, slot = 'scale.data', hm_breaks = c(-3, 0, 10), highlight_pmhc.tcr = T,
             clones_order = clones, clones = clones, show_row_names = T, clean_mat = F, verbose= F )

```

```{r}
pmhc_dist_per_clone(object = obj_sub, clone = '3487_2473', aggr_threshold = 10, slot = 'scale.data')
```

## pMHC-TCR pairing confidence

Each TCR-pMHC interaction recieves a confidence score and p value. Confidence score is calculated as a sum of pmhc's clonotype concordance, log of clone size, subtracted by the noisyness of pMHC. 

\[
\mathbf{Confidence} = \alpha \cdot C + \beta \cdot \log(S + 1) - \gamma \cdot E
\]

\begin{itemize}
  \item \(\alpha, \beta, \gamma\) - weights
  \item \(C\) - clonotype concordance
  \item \(S\) - clone size
  \item \(E\) - pMHC entropy
\end{itemize}

### Permutation Test Procedure

The p-value is obtained by performing two separate permutation tests for the $p\mathrm{MHC}$ value and its stability. These p-values are then combined using Fisher’s method.

For the $p\mathrm{MHC}$ value permutation test, we simulate $n$ subsamples of the same size as the clonotype. We compare the observed $p\mathrm{MHC}$ value (denoted as $V_{\mathrm{obs}}$) within the clone to the distribution of values obtained from simulated clones consisting of random cells.

$$
\text{p-value}_V = \frac{1}{n}\sum_{i=1}^{n}\mathbf{1}\!\left(V_{\mathrm{sim},i} \ge V_{\mathrm{obs}}\right)
$$

For the stability permutation test, we compare the observed stability (denoted as $S_{\mathrm{obs}}$) within the clone to the distribution of stability values obtained from simulated clones.

$$
\text{p-value}_S = \frac{1}{n}\sum_{i=1}^{n}\mathbf{1}\!\left(S_{\mathrm{sim},i} \ge S_{\mathrm{obs}}\right)
$$


```{r}
p_value <- plot_tcr_pmhc_permutation(object = obj_sub, clone = '3487_2473', assay = 'pMHC', slot = 'data')
p_value
```


We can apply some filtering using the parameters, that are calculated during assign_pmhc step.
pvalues filter would use column called pMHC p-value, that uses permitation test for filtering, other options include the use of 
confidence score, and your within-clone assigment test, recorded as pMHC_wclone_pvalue. 

```{r}
filtered_obj <- filter_pmhc(obj_sub, condition = 'pvalues < 0.1') 
pmhc_heatmap(object = filtered_obj, slot = 'scale.data', hm_breaks = c(-3, 0, 10), highlight_pmhc.tcr = T,
             clones_order = clones, clones = clones, show_row_names = T, clean_mat = F, verbose= F )

```

this way loess regression minimises the within clone variability in the pMHC values, forcing outliers to follow the trend, thus denoising the data and making it easy to autocratically assign specificities to each clone. If clone has no signal for any pMHc in your panel, its labeled as negative. The pMHC assignments are stored in `pMHC_classification` column by default

```{r}
obj_sub@meta.data %>%
  select(pMHC_classification, pMHC_confidence, pMHC_pvalues) %>% 
  head(10)
```


You can also record your TCR-pMHC pairs as a separate table using extract_pairs package

```{r}
pairs <- extract_pairs(filtered_obj, 
                       custom_columns = c('BC', 'cdr3_beta', 'cdr3_alpha')) # can add custom columns
head(pairs, 10)
```

```{r session-info, echo=FALSE}
sessionInfo()
```